The weight of the mountains: Constraints on tectonic stress, friction, and fluid pressure in the 2008 Wenchuan earthquake from estimates of topographic loading

### 2014-09-25 revisions/rebuttal.md

Actually, we feel that our manuscript is more similar to the Luttrell et al.
2011 paper because that paper uses a coseismic slip model in the stress
inversion, and because it involves a single event. The Luttrell and Sandwell
paper uses focal mechanisms, which are considered to represent constant coseismic slip over a planar fault plane corresponding to one of the nodal planes of the focal mechanism, and considers
multiple earthquakes. This is a bit more different with respect to our paper.

## R17
> The introduction lists 7 coseismic slip models of the Wenchuan earthquake (line 124), and then the analysis refers to 5 of them (line 213). As the authors noticed, all these models essentially draw from the same set of geodetic data, and all have essentially the same result, and thus all give this study essentially the same result (line 343). Is there a reason these particular models were chosen for analysis? Why not just choose one, or if they truly are different enough to influence these results, why not use them all? Some justification of this choice is warranted.

We used all of the models that we could find that were available (i.e. the slip
model files themselves were published), had proper georeferencing, and seemed
to be of sufficiently high quality. There are many models that did not attempt
a good geometric fit to the available data--e.g. they consist of a single,
planar fault segment--or models with very poor coseismic slip resolution. We
did not feel it wise to include these.

We have added some discussion of the models used, and the criteria for
including them, to Section 2.3, as described in R10. We also reiterate here
that not all of these coseismic slip models relied on the same data,
approximated the fault geometry the same, or used the same inversion strategy
to infer the coseismic slip. As any of these coseismis slip models are just 
*models*, we do feel that basing our conclusions on just one particular model
makes for the most robust study.

### 2015-02-12 revisions/rebuttal2.md

## General criticisms
>1 The grammar and spelling of the manuscript needs to be carefully checked.

In addition to errors caught by reviewers, we have rechecked and corrected
other errors.

>2 The fault geometry needs to be identified early in the manuscript. Right
now, the manuscript does not describe the fault dip anywhere, and this will
cause readers unnecessary difficulty.  These difficulties are particularly
apparent in section 3.

As described below, we have added descriptions of the faults in the text,
and made some figure annotations that we believe will be helpful. We state
here, as below, that the fault geometry is variable within each slip model
and between slip models, such that we cannot state a fault geometry as a
simple strike and dip.

>3 The manuscript confuses (or fails to clearly distinguish between) tractions
and stresses in several places. My experience is that this is common in the
Earth science community. It might be useful in an introductory section to state
that tractions are vectors that act on a particular plane, where as the stress
at a point is a tensor quantity.

We believe that most or all of our readers will know the difference between
tractions and stresses.  However, it is conventional (if slightly improper)
to refer to tractions as stresses, as in 'shear stress' and 'normal stress',
instead of 'shear traction' and 'normal traction'. Whether we are referring
to tensors, vectors, or scalars should be apparent and cause no confusion.

> 4 In several places I had difficulty telling whether the manuscript was
referring to points on the faults or to the fault as a whole. The manuscript
should make the distinction clear.

We have looked over the manuscript for this sort of problem but are not
sure where the reviewer is having trouble, beyond specific comments that are
addressed below.

> 5 The notation for stresses is unconventional, and that makes the manuscript
unnecessarily difficult to read. A more standard notation would be helpful.

We do not agree that it is quite so unconventional, and we have our reasons
for using the notation we chose; we have explained this below, and added some
discussion of this in the manuscript.

> 6 I think it would be better to describe certain distributions or functions
as posterior distributions or posterior functions rather than as “posteriors”
to avoid unintended reactions from some readers.

The term 'posteriors' has been well established in the literature for decades
and is unlikely to cause offense.

> 7 The figures and the text need to be better coordinated so that readers can
easily verify points made in the text when the figures are referred to. Several
of the figures can be modified to make them more useful to the reader. These
modifications should not be time-consuming to make.

We have made some minor modifications to figures.

> 8 A key modeling assumption seems to be that fault slip is not affected by
stress changes accompanying the fault rupture, but this assumption is not
stated. I think it should be.

We clearly state that our results are predicated upon the Wallace-Bott
assumption that fault slip occurs in the direction of maximum shear stress.

> 9 The comparison of the Beichuan fault to an optimally oriented fault needs
to be clarified. This seems to be a key point, but the discussion of this point
is not sufficiently clear.

We have reworked some of this, based on the comments.  However, it does seem
that the reviewer missed the key point that friction on the Beichuan fault
is less than friction on the hypothetical optimally-oriented faults, though
this is quite clear in the text.

## Specific Comments

### Abstract

**CA.1**

> Do they mean the whole fault was verging on failure at the time of the
earthquake, or only part of it?

We have changed the relevant sentence to 'We further estimate the static
friction and pore fluid pressure assuming that the fault was*, on average,* at
Mohr-Coulomb failure at the time of the Wenchuan earthquake.

**CA.2**

> Do they account for the topographic perturbation of the tectonic stresses in
coming up with this estimate? See for example Savage and Swolfs [1986].

We do not, because the method of Savage and Swolfs (as well as Liu and Zoback,
from whom we take our methods for topographic stress calculations) assume
an *a priori* tectonic stress value that is a constant. However, we assume that
tectonic stress is a tensor field that increases with depth; and we do not know
it *a priori*. Therefore, it is difficult to determine what the perturbation
to tectonic stresses is.

We have explained this more as part of the previous revisions.

### 1. Introduction

**C1.1**

> The only omission in literature coverage I am aware of is the work by
McGovern on topographic stresses associated with volcanoes. I think some of
McGovern’s papers from the 1990’s should be cited.

We have looked over these references, which are largely about volcanic
deformation on other planets, and have not found a good place to cite them,
without making a separate paragraph for discussions of topographic stresses in
volcanoes. This is far from our topic (the interaction of topographic stresses
and tectonics), and we did not read these papers during the work presented in
this paper, so we don't feel it appropriate to cite them.

**C1.2**

> This is the section of the manuscript where both the strike and dip of the
key faults should be stated.

We have added a paragraph and a half of description of the faults to the
introduction in the previous round of revisions.

**C1.3**

> (40-41) Referring to the Qur’an (or any other religious text) as one might
refer to a peer- reviewed scientific article is highly inappropriate. This
reference must be eliminated.

This reference has been eliminated, with the note that it is unfortunate that
we put up these sorts of walls in scholarship. The history of ideas shouldn't
be segregated so, and we did not refer to it 'as one might a peer-reviewed
scientific article'.

### 2. Topographic stresses on the Longmen Shan faults

**C2.1**

> This section came across well. I do think a description of the mechanics
behind equations 5 and 6 would be helpful though.

**C2.2**

> (169-170) The authors should be a bit more clear about some of the
limitations of the method of Liu and Zoback [1992]. That method cannot be used
to accurately calculate the topographically-induced stresses at shallow depths.
I think the method of Liu and Zoback is fine to use for the problem at hand,
but the method is not exact, and it does have its limitations. Those
limitations should be pointed out.

We change the sentence introducing the methods to "They show that the
topographic stress tensor field *beneath (but not within) topography* can be
determined by a convolution of topographic loading functions with Green's
functions describing the stresses in an elastic halfspace due to a point load
at the surface." (changes in italics). Additionally, in Section 2.2 on the
numerical implementation of the methods, we mention that our calculations only
apply to depths below 851 m.

**C2.3**

> (172) Why is M used to designate stress rather than $\sigma$, as is
conventional?  Some people might think M is for Moment. I strongly recommend
using σ for stresses.

Capital letters are quite commonly used to denote matrices, tensors, and
tensor fields. *M* is clearly defined in the manuscript.

See the response to Comment 3.9b, where we add text to the manuscript
explaining our notational usage.

**C2.4**

> (174) The line immediately after all the equations is indented and shouldn't
be. The indentations are supposed to signify the start of a new paragraph.

These are fixed.

**C2.5**

> (206-208) The open-source package is nice to make available to the community.

Yes, hopefully this will be useful for others in the future.

### 3. Results of topographic stress calculations on the Wenchuan faults

**C3.1**

> The manuscript needs to present figures and refer the reader to them to
verify the statements made in this section about the stresses. Right now
I can’t tell where to look to do this in several places in this section. Has
the manuscript omitted a figure?

No, the manuscript isn't missing any figures. Our results for this section
include each of the six stress tensor components, as well as the resolved
tractions ($\tau^M$, $\tau_d^M$, $\tau^M_s$, $\sigma_n^M$) for each of the six
coseismic slip models we've used. This would be 60 subfigures, clearly too many
to include as figures in the manuscript, so we have selected the most
illustrative results. All of the results are publicly available as .csv files
in the GitHub repository for project, and readers are welcome to download and
analyze these datasets if they wish to verify our claims, or otherwise inspect
the results. We have added a sentence informing readers of this to the end of
the first paragraph in Section 3:

"(Results not shown in figures are available as .csv files at
http://github.com/cossatot/wenchuan_topo_stress/)."

Additionally, in response to comments in the earlier round of revisions, we
have added vertical stresses to Figures 3 and 9.

**C3.2**

> Also, the manuscript needs a figure showing the model geometry. Also, I have
difficulty with the non-standard notation.

As mentioned in the response to Comment 1.2, we have added a descriptions of
the coseismic slip models used, and on the general geometry of the relevant
faults, to Section 2.3, as part of the previous rounds of revisions.

Additionally, as is evident in Figure 4, there is considerable variation in the
fault geometries assumed or inverted for by different researchers, so it is not
possible to accurately show a figure with model geometry, because there is no
real consensus on it. (To be clear, shallower than about 8 km depth, all of the
slip models we've used are quite similar in geometry and location, but below
this depth they start to diverge, as some models assume planar geometry and
others assume listric geometry, and so forth).

We have added some annotation of the dips and depths of various parts of the
slip models shown on Figure 4, in order to help readers.

The issue with notation is addressed in the response to Comment 3.9b.

**C3.3**

> Much of this section needs to be reworked so it is clear to the reader.

Specific criticisms of clarity are addressed in the following
comments and responses.

**C3.4a**

> (220) What is the purpose of the superscript M here for the traction
components? If does not seem to serve a purpose. Is it to designate tractions
due to topographic loads? If so, I recommend the authors be more explicit about
that.

Yes, the superscript M denotes the topographic origin for stresses. See the
response to Comment 3.9b.

**C3.4b**

> (220) See my comment on line 243 regarding the notation.

See our response to Comment 3.9b.

**C3.5a**

> (222-230) I can’t tell where to look to verify the statements about
stresses. Which figure, for example, shows Myz, the vertical normal stress?
Which figure shows Myz, the horizontal shear stress on a vertical plane that
strikes north?

See our response to Comment 3.1.

**C3.5b**

> (222-230) At what depth or location is the first sentence of the paragraph
referring to?

This sentence ("Topographic stresses on the Wenchuan faults are on the order
1-10s MPa") refers to the entirety of the faults, although it is a general
sentence and there are a few points on the faults where the stresses are below
this range.

**C3.6**

> (225-226)  Which figure shows the vertical normal stress? I can’t see a plot
in Fig. 3 or Fig. 4 that shows the vertical normal stress.

See our response to Comment 3.1.

**C3.7**

> (240-241) Which figure should the reader look at to see the faults in the
model?

We have added a reference to Figure 4b and c.

**C3.8**

> (242) Where are the locations being referred to here? What are their x,y,z
coordinates (or latitude, longitude, and depth)?

The Pengguan fault is shown on the maps (Figures 1, 3, 9). The flats in the
Beichuan fault are at the base of the southwestern part of the fault, as is
evident in Figure 4, which shows some of the fault models. We think that giving
'x,y,z' or other spatial coordinates would be quite unhelpful (particularly for
curved surfaces tens of kilometers in breadth), so we will not do this.  If
readers need to know this information, they can look at the .csv files on
GitHub or consult the sources for the slip models.

**C3.9a**

> (243) What is the dip of the fault in question?

This varies in the different fault models considered. If the reader wants to
know these specifics, he or she is welcome to read the papers that have
produced the fault models used.

**C3.9b**

> (243) The notation here, where M (with
double subscripts) refers to a stress, $\sigma$ (with a single subscript) to a
stress (but it would be more appropriately called a traction), and $\tau$ (with
a single subscript) to a stress (but again it would be more appropriately
called a traction) is very confusing to me. I strongly recommend using a
convention like so:

>$\sigma_{ij}$ = stress

>$\tau_n$= normal traction on a fault

>$\tau_s$ = strike-slip traction on a fault

>$\tau_d$ = dip-slip traction on a fault

>This is much more in keeping with standard usage and would be easier for
(most) readers to follow I think.

I do not think I have ever seen $\tau_n$ to refer to normal stress/traction;
we use $\tau$ for shear stress (tractions), which is quite standard.  We also
use $\sigma_n$ as normal stress (traction) which is standard. We use *M* to
denote the topographic stress tensor or tensor field, and it is clearly defined
as such. Capital letters are extremely commonly used to denote arrays,
matrices, tensors and tensor fields; this is not only a structural geology
convention but spans the science, mathematics and engineering disciplines.

We considered adopting the reviewer's suggestions (outside of using $\tau_n$
for normal stress) but in the end we do not use $\sigma$ to represent a stress
tensor or tensor field because we would have to exponentiate it to, for
example, $\sigma^{M^B}$ or $\sigma^{M,B}$ when referring to the stress tensor
field resulting from the convolution of topography with the Boussinesq Green's
functions. This is unwieldy. It would also be quite awkward to refer to the
normalized (dimensionless) tectonic stress field as $\sigma^{T^{\prime}}$ even
though it does not have values in Pa, or we could keep it as $T^\prime$ which
could further confuse readers as it is then one more step removed from *T*.

We have added an explanation of our notation to the first paragraph of Section
2.1:  "Note that in this work, we use capital letters (e.g., $M$, $T$) to
denote tensors and tensor fields (depending on context, which should be clear)
and Greek letters ($\tau$, $\sigma$) to denote stress components projected or
resolved on planes, which are more properly called tractions. We use
superscripts on these symbols to denote the origin of the stresses, and
subscripts to denote components of these tensors or stresses or tractions."

**C3.10**

> (244-245) Which figure should the reader look at to verify the statement
here?

We disagree that a figure is necessary, or particularly helpful, to verify a
claim that vertical stresses resolve more as normal stress than shear stress
on a plane with low dip. This should be apparent to readers who have taken a
course in Structural Geology or who have a continuum mechanics background,
which surely includes the vast majority of our readers. Readers without this
background will unfortunately have a quite difficult time with the paper, but
they are not our intended audience.

**C3.11**

> (249-251) I don't see where to look to verify the nature of the slip and the
tractions on Figure 4.  The blue background color of the slip magnitude figure
(upper left) looks very uniform, so I can't tell where the slip is high; also
note that the contour lines are not labeled with magnitudes. In the normal
stress plot in the upper right the red background color is very uniform, so
I can't tell where the normal traction is high (and the contour lines are not
labeled). This means I don't know how to compare the plots.

The colors shown are not simple 'background colors' but correspond to the
magnitudes of the stress component in question. Where they are uniform,
stresses are uniform. Labeling the contours would add unneccesary clutter to
the figures, making them harder to read. Nonetheless, the relevant comparisons
(for example, between high normal stress and low slip) are still quite evident
to us, both on the screen and in print, and other viewers have not had problems
with this. Maybe the reviewer has a monitor or printer calibration issue.

**C3.12**

> Where are these other fault patches? How big are they? Can the authors point
the reader to a figure to show these patches?

The 'other' fault patches are those that are not the ones described (i.e. fault
segments that are not the southwestern portion of the Beichuan fault). Figure
4 (referenced immediately before the sentence in question) illustrates where
the observed correlation between slip and normal stress is strong and where
it is not.

### 4. Calculations of tectonic stress, fault friction and pore fluid pressure

**C4.1**

> This section reads fairly well, although a fuller description of the Bayesian
statistics would have helped me.

We believe that we have explained the mathematics involved in our calculations
(Bayes' rule, the misfit equation, and the likelihood equation, which are
equations 10-12 in the original submission and equations 11-13 in the revised
submission). These are uncomplicated expressions. Unfortunately, though
Bayesian statistics are intuitive in practice and a formalized version of our
thought processes in many instances (revising ideas or beliefs based on
evidence), there is some sort of barrier for many people in grasping it
initially. For the first author, this was due to jargon, so we have tried to
define and explain each of the terms. We have also cited two excellent texts
[Tarantola, 2005; Mosegaard and Tarantola, 1995] that explain Bayesian
statistics in more detail.

**C4.2**

> (256-257) The statement here should also account for cohesion, although the
frictional resistance could be much larger than any intrinsic cohesion

This is true; we have added "and assuming that rock strength (cohesion) is
negligible" to the sentence describing the Mohr-Coulomb failure envelope.

**C4.3**

> (258) I think the authors mean when the earthquake initiated, not the fault.
The fault is presumably millions of years old.

Yes, this is what we meant.  We have replaced 'fault' with 'earthquake' in this
sentence.

**C4.4a**

> (265) The phrase "posteriors of T" could have unanticipated (and undesirable)
reactions from some readers. I suggest using the alternative phrase "these
functions".

The use of the word 'posterior' in statistics is well-established, and surely
won't incite any more unintended reactions or schoolboy jokes than 'thrusting'.
We advise readers who blush at the term 'posterior' to avoid television and
the internet at all costs.

**C4.4b**

> (265) Also, what is "T"?

*T* has been defined in this sentence as part of the previous revisions.

**C4.5**

> (274-275) The grammar and punctuation here need to be corrected.

We have fixed these problems as part of the previous revisions.

**C4.6**

> (276-287) Owing to my poor background in Bayesian statistics, I cannot
comment knowledgeably on the technical aspects of this paragraph except for one
point.  The authors should clarify whether the are evaluating the direction of
slip based on the conditions before the fault has begun to slip or as the fault
slips and the stress field around the slipped patch is evolving.

The direction of slip is given by the coseismic slip models, which are the net
slip from the earthquake; this is described in Section 4.2.  A consideration
of rake variability during the rupture would be excellent but unfortunately
we do not have the data to do this.

**C4.7**

> (290) The authors need to clarify what orientations they are referring to
here.

We have clarified the orientations as stress orientations.

**C4.8**

> (291-292) The statement here is true, but only because the lithostatic
stresses are defined to have a horizontal components equal to the vertical
components. The <u>definition</u> of the tensor L thus plays a key role in the
results.

Yes it does, but we have been explicit in our definition of *L* (section 4.1).

**C4.9**

> (293-294) Is the maximum tectonic principal stress being referred to here? If
so, I think it would be a good idea to state it here, even though the topic
sentence of the paragraph deals with principal tectonic stresses. It is very
important that the critical thoughts in the manuscript at this point be crystal
clear to the reader.

We have added 'tectonic' to 'maximum principal *tectonic* stress' and 'minimum
principal *tectonic* stress' to clarify this in these sentences.

### 5. T, $\mu$, $\phi$ results

**C5.1**

> Owing to my poor background in Bayesian statistics, this section was
difficult for me to understand. Other readers might be in the same situation.
I think the section would be substantially improved if it concluded with a new
brief paragraph that was written “in plain English” that gave the key points
the reader should take away from this section.

We do not find our descriptions of the results to require much of a knowledge
of Bayesian statistics.  It might be that the notation (which is the standard
notation of conditional probabilities) is confusing, but this is defined in
the preceeding sections.

**C5.2**

> In spite of my shortcomings in Bayesian statistics, I am pretty sure the key
point that the reader should take away from this section is not provided by the
last sentence in the section (“On the other hand, it is not similarly sparse on
the low-μ side, suggesting that low values for μ are more robust.”).

The placement of the key points of a section to be in the last sentence of that
section is a stylistic preference, and not one that we share (why not put the
important information towards the top of the paragraphs rather than at the
end?). There is not really a 'key point' here per se; this is a description of
the results. The important results are discussed as such.

**C5.3**

> (328-330) After this sentence, say what the significance of the peak in
T'(min) at 0.2 is, as was done in lines 326-328 for T'(max).
